From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-30 Pi/paper/equivalence.tex

The translations between the languages are defined separately on types, 1-combinators, and 2-combinators. Following the
terminology of Normalisation by Evaluation, the translations from the left to the right, going from the syntax towards
the semantics, are called $\evalt$ and the translations the other way are called $\quotet$.

@@

\section{Normalisation of Reversible Circuits}
\label{sec:applications}

Using our semantics, we can normalise, synthesise, prove equivalence, and generally reason about~$\PiLang$ programs. The
two key definitions are presented below.

\begin{definition}[Normalisation of $\PiLang$ programs]
  \begin{gather*}
    \begin{aligned}
      \normt_{0} & : U \to \UPlus                                            \\
      \normt_{0} & = \quotep_{0} \comp \quoteh_{0} \comp \gdenot{\blank}_{0}
    \end{aligned}
    \qquad
    \begin{aligned}
      \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \iso \normt_{0}(Y)     \\
      \normt_{1} & = \quotep_{1} \comp \quoteh_{1} \comp \gdenot{\blank}_{1}
    \end{aligned}
  \end{gather*}
\end{definition}

\noindent Normalisation involves translating $\PiLang$ programs to $\PiHatLang$, computing a permutation, and quoting
back to $\PiPlusLang$. Note that the normalisation happens in the step from $\PiHatLang$ to $\UFin$ and back to
$\PiHatLang$ and that normalisation also provides a decision procedure for program equivalence. Synthesis happens by
quoting permutations. More general, user-guided, reasoning can be done using the sound and complete level-2
combinators to rewrite $\PiLang$ programs.

Recall the specification of reversible disjunction from \cref{sec:examples}, the two Qiskit circuits for implementing
it, and the corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The type $\mathbb{B}\ 3$
normalises to $\mathbb{8}+ = \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} +
(\mathbb{1} + \mathbb{0})))))))$, and normal forms for both circuits compute to the following, establishing their equivalence:

\medskip
\resetnormtwo{}

Instead of manually producing $\Pi$-programs to implement the reversible disjunction specification, it is possible to
simply quote the desired permutation (using a function $\term{lookup}$ to read a tabulated version for readability):

\medskip
\resetperm{}

\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Quoting this
permutation generates the same normalised program which can then be composed with a map from $\PiLang$ to
$\PiPlusLang$ to produce a program matching the desired structured types.

% \paragraph*{Program synthesis.} The NbE process embodies a quoting mechanism that synthesizes programs from
% permutations. Indeed, instead of writing a program for \Afun{reset 2}, one could simply specify the desired permutation
% as:

%%\begin{minipage}{.65\textwidth}
%%  \PiRESET{}
%%\end{minipage}
%%\begin{minipage}{.30\textwidth}
%%  \begin{center}
%%  \resizebox{0.5\textwidth}{!}{\input{diagrams/reset.tikz}}
%%  \end{center}
%%\end{minipage}

%% \resettwo{}

% \noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
% provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
% reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the introduction.

% In principle, the normalised program can be produced following two strategies: (i) by repeatedly applying the rewrite
% rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
% using a NbE process that evaluates the program to a permutation on a finite set of 8
% elements and reifies that permutation back to a program. The key idea of the NbE process is a systematic way to express
% permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
% left is compiled to the sequence of four adjacent transpositions on the right:

% \note{Motivation: There are two reversible circuits which describe the following permutation. They can be shown to be
%   equal using the 2-combinators.}

% \[
%   \begin{tikzpicture}
%     \begin{knot}[clip width=5]
%       \filldraw (0,5) circle (2pt) node[above] {0};
%       \filldraw (1,5) circle (2pt) node[above] {1};
%       \filldraw (2,5) circle (2pt) node[above] {2};
%       \filldraw (3,5) circle (2pt) node[above] {3};
%       \filldraw (4,5) circle (2pt) node[above] {4};
%       \filldraw (0,0) circle (2pt) node[below] {1};
%       \filldraw (1,0) circle (2pt) node[below] {4};
%       \filldraw (2,0) circle (2pt) node[below] {0};
%       \filldraw (3,0) circle (2pt) node[below] {3};
%       \filldraw (4,0) circle (2pt) node[below] {2};
%       \strand (0,5) .. controls (0.5,0.5) and (1.5,3.5) .. (2,0);
%       \strand (1,5) .. controls (0.75,0.5) and (0.25,3.5) .. (0,0);
%       \strand (2,5) .. controls (2.5,2.5) and (3.5,1.5) .. (4,0);
%       \strand (3,5) .. controls (4.5,2.5) and (4,1.5) .. (3,0);
%       \strand (4,5) .. controls (3.5,2.5) and (1.5,2.5) .. (1,0);
%       \flipcrossings{4,5};
%     \end{knot}
%   \end{tikzpicture}
% \]

% \paragraph*{Program equivalence.} The permutation above reveals another way to think about the desired program: it is a
% special addition circuit that keeps 0 and 4 fixed but otherwise adds 4 modulo 8 to its input. From this specification,
% one can use a standard synthesis algorithm for reversible circuits~\cite{10.1145/775832.775915} to generate the
% following program:

% \adder{}

% \noindent The \Afun{adder3} program looks nothing like the original \Afun{reset 2} program and yet they both have the
% same normal form thus establishing their equivalence. The reader can check that this circuit is the same as
% \verb|reversibleOr1| from the introduction.

% Now imagine we want to write the following reversible function:

% 0 -> 0
% 8 -> 8
% n -> n + 8 `mod` 16

>>>>

The translations between the languages are defined separately on types, 1-combinators, and 2-combinators. Following the
terminology of Normalisation by Evaluation (NbE), the translations from the left to the right, going from the syntax
towards the semantics, are called $\evalt$ and the translations the other way are called $\quotet$.

@@

### 2021-10-30 Pi/paper/introduction.tex

\paragraph{Outline and Contributions} Our main result is a proof of the soundness and completeness of $\PiLang$ with
respect to its semantics in the weak symmetric rig groupoid of finite sets and permutations. We state our result as a
Curry-Howard-Lambek correspondence for Reversible Logic, Reversible Programming Languages, and Symmetric Rig Groupoids,
using which we can build a toolbox of technical devices for reasoning about reversible circuits.

>>>>

\paragraph{Outline and Contributions} Our main result is a proof of the soundness and completeness of $\PiLang$ with
respect to its semantics in the weak symmetric rig groupoid of finite sets and permutations, formalised in HoTT/UF. We
state our result as a Curry-Howard-Lambek correspondence for Reversible Logic, Reversible Programming Languages, and
Symmetric Rig Groupoids, using which we can build a toolbox of technical devices for reasoning about reversible
circuits.
